```{r, echo=FALSE}
source(file = "setup.R")
```

# Kruskal-Wallis test

## Objectives
:::objectives
**Questions**

- How do I analyse multiple samples of continuous data if the data are not normally distributed?
- What is a Kruskal-Wallis test?
- How do I check for differences between groups?

**Objectives**

- Be able to perform an Kruskal-Wallis test in R
- Understand the output of the test and evaluate the assumptions
- Be able to perform post-hoc testing after a Kruskal-Wallis test
:::

## Purpose and aim
The Kruskal-Wallis one-way analysis of variance test is an analogue of ANOVA that can be used when the assumption of normality cannot be met. In this way it is an extension of the Mann-Whitney test for two groups.

## Section commands
New commands used in this section:

| Function| Description|
|:- |:- |
|`kruskal_test()`| Performs the Kruskal-Wallis test |
|`dunn_test()`| Performs Dunn's test |

## Data and hypotheses
For example, suppose a behavioural ecologist records the rate at which [spider monkeys](https://en.wikipedia.org/wiki/Spider_monkey) behaved aggressively towards one another as a function of closely related the two monkeys are. The familiarity of the two monkeys involved in each interaction is classified as `high`, `low` or `none.` We want to test if the data support the hypothesis that aggression rates differ according to strength of relatedness. We form the following null and alternative hypotheses:

-	$H_0$: The median aggression rates for all types of familiarity are the same
-	$H_1$: The median aggression rates are not all equal

We will use a Kruskal-Wallis test to check this.

The data are stored in the file `data/raw/CS2-spidermonkey.csv`.

First we read the data in:

```{r cs2-kruskal-data, message=FALSE, warning=FALSE}
spidermonkey <- read_csv("data/tidy/CS2-spidermonkey.csv")
```

## Summarise and visualise

```{r cs2-kruskal-sumvisual}
# look at the data
spidermonkey

# summarise the data
spidermonkey %>% 
  select(-id) %>% 
  group_by(familiarity) %>% 
  get_summary_stats(type = "common")

# create boxplot
spidermonkey %>% 
  ggplot(aes(x = familiarity, y = aggression)) +
  geom_boxplot()
```

The data appear to show a very significant difference in aggression rates between the three types of familiarity. We would probably expect a reasonably significant result here.

## Assumptions
To use the Kruskal-Wallis test we have to make three assumptions:

1.	The parent distributions from which the samples are drawn have the same shape (if they’re normal then we should use a one-way ANOVA)
2.	Each data point in the samples is independent of the others
3.	The parent distributions should have the same variance

Independence we’ll ignore as usual. Similar shape is best assessed from the earlier visualisation of the data. That means that we only need to check equality of variance.

### Equality of variance

We test for equality of variance using Levene’s test (since we can’t assume normal parent distributions which rules out Bartlett’s test).

```{r cs2-kruskal-levene}
# perform Levene's test
spidermonkey %>% 
  levene_test(aggression ~ familiarity)
```

The relevant p-value is given in the `p` column (0.893). As it is quite large we see that each group do appear to have the same variance.

There is also a warning about `group coerced to factor`. There is no need to worry about this - Levene's test needs to compare different groups and because `aggression` is encoded as a numeric value, it converts it to a categorical one before running the test.

## Implement test
Perform a Kruskal-Wallis test on the data:

```{r cs2-kruskal-test, results='hide'}
# implement Kruskal-Wallis test
spidermonkey %>% 
  kruskal_test(aggression ~ familiarity)

kruskal.test(aggression ~ familiarity, data = spidermonkey)
```

-	The `kruskal_test()` takes the formula in the following format: `variable ~ category`

## Interpret output and report results
This is the output that you should now see in the console window:

```{r cs2-kruskal-test-result, echo=FALSE}
# implement Kruskal-Wallis test
spidermonkey %>% 
  kruskal_test(aggression ~ familiarity)

kruskal.test(aggression ~ familiarity, data = spidermonkey)
```

The p-value is given in the `p` column. This shows us the probability of getting samples such as ours if the null hypothesis were actually true.

Since the p-value is very small (much smaller than the standard significance level of 0.05) we can say "that it is very unlikely that these three samples came from the same parent distribution and as such we can reject our null hypothesis" and state that:

> A one-way Kruskal-Wallis rank sum test showed that aggression rates between spidermonkeys depends upon the degree of familiarity between them (KW = 13.597, df = 2, p = 0.0011).

<br />

## Post-hoc testing (Dunn's test)
The equivalent of Tukey’s range test for non-normal data is **Dunn’s test**.

Dunn's test is used to check for significant differences in group medians:

```{r cs2-dunn-test, results='hide'}
# perform Dunn's test
spidermonkey %>% 
  dunn_test(aggression ~ familiarity)
```

This will give the following output:

```{r cs2-dunn-test-result, echo=FALSE}
# perform Dunn's test
spidermonkey %>% 
  dunn_test(aggression ~ familiarity)
```

The `dunn_test()` function performs a Kruskal-Wallis test on the data, followed by a post-hoc pairwise multiple comparison.

The comparison between the pairs of groups is reported in the table at the bottom. Each row contains a single comparison. We are interested in the `p` and `p.adj` columns, which contain the the p-values that we want. This table shows that there isn’t a significant difference between the high and low groups, as the p-value (0.1598) is too high. The other two comparisons between the high familiarity and no familiarity groups and between the low and no groups are significant though.

The `dunn_test()` function has several arguments, of which the `p.adjust.method` is likely to be of interest. Here you can define which method needs to be used to account for multiple comparisons. The default is `"holm"`. We'll cover more about this in the chapter on [Power analysis](#cs6-intro).

## Exercise: Lobster weight
:::exercise
Kruskal-Wallis and Dunn's test on lobster data

Perform a Kruskal-Wallis test and do a post-hoc test on the `lobster` data set.

<details><summary>Answer</summary>

### Hypotheses

- $H_0$ : all medians are equal
- $H_1$ : not all medians are equal

### Import data, summarise and visualise
```{r cs2-kruskal-lobster-read, echo=FALSE, message=FALSE, warning=FALSE}
# load the data in the background
lobsters <- read_csv("data/tidy/CS2-lobsters.csv")
```

All done previously.

### Assumptions

From before, since the data are normal enough they are definitely similar enough for a Kruskal-Wallis test and they do all have equality of variance from out assessment of the diagnostic plots. For completeness though we will look at Levene's test

```{r cs2-levene-lobster, warning=FALSE}
lobsters %>% 
  levene_test(weight ~ diet)
```

Given that the p-value is so high, this again agrees with our previous assessment that the equality of variance assumption is well met. Rock on.

### Kruskal-Wallis test

```{r cs2-kruskal-lobster}
# implement Kruskal-Wallis test
lobsters %>% 
  kruskal_test(weight ~ diet)
```

> A Kruskal-Wallis test indicated that the median weight of juvenile lobsters did not differ significantly between diets (KW = 3.26, df = 2, p = 0.20).

<br />

### Post-hoc Dunn's test

Although rather unnecessary (and likely unwanted, since we don't want to be p-hacking), because we did not detect any significant differences between diets, we can perform the non-parametric equivalent of Tukey's range test: Dunn's test.

```{r cs2-dunn-lobster}
# perform Dunn's test
lobsters %>% 
  dunn_test(weight ~ diet)
```

We can see that none of the comparisons are significant, either based on the uncorrected p-values (`p`) or the p-values adjusted for multiple comparisons (`p.adj`). This is consistent with what we found previously.

</details>
:::

## Key points

:::keypoints
- We use Kruskal-Wallis test to see if there is a difference in medians between multiple continuous variables
- In R we first define a linear model with `lm()`, using the format `response ~ predictor`
- Next, we perform a Kruskal-Wallis test on the linear model with `kruskal_test()`
- We assume parent distributions have the same shape; each data point is independent and the parent distributions have the same variance
- We test for equality of variance using `levene_test()`
- Post-hoc testing to check for significant differences in the group medians is done with `dunn_test()`
:::
