```{r, echo=FALSE}
source(file = "setup.R")
```

# Model comparisons

## Objectives
:::objectives
**Questions**

- How do I compare linear models?
- How do decide which one is the "best" model?

**Objectives**

- Be able to compare models using the Akaike Information Criterion (AIC)
- Use AIC in the context of Backwards Stepwise Elimination in R
:::

## Purpose and aim
In the previous example we used a single dataset and fitted five linear models to it depending on which predictor variables we used. Whilst this was fun (seriously, what else would you be doing right now?) it seems that there should be a "better way". Well, thankfully there is! In fact there a several methods that can be used to compare different models in order to help identify "the best" model. More specifically, we can determine if a full model (which uses all available predictor variables and interactions) is necessary to appropriately describe the dependent variable, or whether we can throw away some of the terms (e.g. an interaction term) because they don’t offer any useful predictive power.

Here we will use the **Akaike Information Criterion** in order to compare different models.

## Section commands
New commands in this section:

| Function| Description|
|:- |:- |
|`extractAIC()`| Extract the Akaike Information Criterion from a fitted model |
|`step()` | Performs backwards stepwise elimination on a model |

## Data and hypotheses
This section uses the `data/tidy/CS5-Ladybird.csv` data set. This data set comprises of 20 observations of three variables (one dependent and two predictor). This records the clutch size (`eggs`) in a species of ladybird alongside two potential predictor variables; the mass of the female (`weight`), and the colour of the male (`male`) which is a categorical variable.

## Backwards Stepwise Elimination
First, load the data and store it in an object called `ladybird`. Then visualise the data.

```{r, message=FALSE, warning=FALSE}
ladybird <- read_csv("data/tidy/CS5-Ladybird.csv")
```

```{r}
# visualise the data
ladybird %>% 
  ggplot(aes(x = weight, y = eggs,
             colour = male)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2")
```

### Comparing models with AIC (step 1)
First, we construct the full linear model:

```{r, results='hide'}
# define the full model
lm_full <- lm(eggs ~ weight * male,
              data = ladybird)

# view the model summary
summary(lm_full)
```

Now we construct a reduced model (i.e. the next simplest model) which doesn’t have interactions:

```{r, results='hide'}
# define the model
lm_red <- lm(eggs ~ weight + male,
             data = ladybird)

# view the model summary
summary(lm_red)

```

To compare the two models we simply use the command `extractAIC()` on each model.

```{r}
extractAIC(lm_full)
extractAIC(lm_red)
```

For each line the first number tells you how many parameters are in your model and the second number tells you the AIC score for that model. Here we can see that the full model has 4 parameters (the intercept, the coefficient for the continuous variable `weight`, the coefficient for the categorical variable `male`  and a coefficient for the interaction term `weight:male`) and an AIC score of 41.3 (1dp). The reduced model has a lower AIC score of 40.4 (1dp) with only 3 parameters (since we’ve dropped the interaction term). There are different ways of interpreting AIC scores but the most widely used interpretation says that:

* if the difference between two AIC scores is **greater than 2** then the model with the **smallest AIC score is more supported** than the model with the higher AIC score
* if the difference between the two models’ AIC scores is **less than 2** then both models are **equally well supported**

This choice of language (supported vs significant) is deliberate and there are areas of statistics where AIC scores are used differently from the way we are going to use them here (ask if you want a bit of philosophical ramble from me). However, in this situation we will use the AIC scores to decide whether our reduced model is at least as good as the full model. Here since the difference in AIC scores is less than 2, we can say that dropping the interaction term has left us with a model that is both simpler (fewer terms) and as least as good (AIC score) as the full model. As such our reduced model `eggs ~ weight + male` is designated our current _working minimal model_.

### Comparing models with AIC (step 2)
Next, we see which of the remaining terms can be dropped. We will look at the models where we have dropped both `male` and `weight` (i.e. `eggs ~ weight` and `eggs ~ male`) and compare their AIC values with the AIC of our current minimal model (`eggs ~ weight + male`). If the AIC values of at least one of our new reduced models is lower (or at least no more than 2 greater) than the AIC of our current minimal model, then we can drop the relevant term and get ourselves a new minimal model. If we find ourselves in a situation where we can drop more than one term we will drop the term that gives us the model with the lowest AIC.

Drop the variable `weight` and examine the AIC:

```{r}
# define the model
lm_male <- lm(eggs ~ male,
              data = ladybird)

# extract the AIC
extractAIC(lm_male)

```

Drop the variable `male` and examine the AIC:

```{r}
# define the model
lm_weight <- lm(eggs ~ weight,
                data = ladybird)

# extract the AIC
extractAIC(lm_weight)

```

Considering both outputs together and comparing with the AIC of our current minimal model (40.4) we can see that dropping `male` has decreased the AIC further to 38.8, whereas dropping `weight` has actually increased the AIC to 60.0 and thus worsened the model quality.

Hence we can drop `male` and our new minimal model is `eggs ~ weight`.

### Comparing models with AIC (step 3)
Our final comparison is to drop the variable `weight` and compare this simple model with a null model (`eggs ~ 1`), which assumes that the brood size is constant across all parameters.

Drop the variable `weight` and see if that has an effect:

```{r}
# define the model
lm_null <- lm(eggs ~ 1,
              data = ladybird)

# extract the AIC
extractAIC(lm_null)

```

The AIC of our null model is quite a bit larger than that of our current minimal model `eggs ~ weight` and so we conclude that `weight` is important. As such our minimal model is `eggs ~ weight`.

So, in summary, we could conclude that:

> Female size is a useful predictor of clutch size, but male type is not so important.

At this stage we can analyse the minimal linear (`lm.weight`) model using the `anova()` function, and we should consider the diagnostic plots by using the `plot(lm.weight)` command.

## Notes on Backwards Stepwise Elimination
This method of finding a minimal model by starting with a full model and removing variables is called backward stepwise elimination. Although regularly practised in data analysis, there is increasing criticism of this approach, with calls for it to be avoided entirely.

Why have we made you work through this procedure then? Given their prevalence in academic papers, it is very useful to be aware of these procedures and to know that there are issues with them. In other situations, using AIC for model comparisons are justified and you will come across them regularly. Additionally, there may be situations where you feel there are good reasons to drop a parameter from your model – using this technique you can justify that this doesn’t affect the model fit. Taken together, using backwards stepwise elimination for model comparison is still a useful technique.

:::note
Performing backwards stepwise elimination manually can be quite tedious. Thankfully R acknowledges this and there is a single inbuilt function called `step()` that can perform all of the necessary steps for you using AIC.

```{r, echo=FALSE}
# define the full model
lm_full <- lm(eggs ~ weight * male,
              data = ladybird)

# perform backwards stepwise elimination
step(lm_full)
```

This will perform a full backwards stepwise elimination process and will find the minimal model for you. The output should be familiar to you but ask a demonstrator if you have any questions.

Yes, I could have told you this earlier, but where’s the fun in that? (it is also useful for you to understand the steps behind the technique I suppose...)
:::

## Exercise: BSE
:::exercise
BSE on `trees` and `airpoll`

Use the internal dataset `trees` and the `airpoll` dataset from earlier.

* Perform a backwards stepwise elimination on both of these datasets and discover the minimal model using AIC.

NB: if an interaction term is significant then any main factor that is part of the interaction term cannot be dropped from the model.

* If you’re feeling up for it attempt a backwards stepwise elimination process on the internal `CO2` dataset. This data frame has 1 dependent variable (`uptake`) and 4 predictor variables (`Plant`, `Type`, `Treatment`, `conc`). Unfortunately, the dataset does not contain enough data to construct a full linear model using all 4 predictor variables (with all of the interactions), so ignore the `Plant` variable and take `uptake ~ Type + Treatment + conc + Type:Treatment + Type:conc +  Treatment:conc + Type:Treatment:conc` as your full model.

<details><summary>Answer</summary>

This is relatively straightforward using the `step()` function.

We need to first construct the full linear model and then simply pass this linear model object to the step function and R does the rest.

### `trees` dataset

We construct a full linear model with both `Height`, `Girth` and the interaction between them and then we run the `step()` function:

```{r}
# define the full model
lm_trees <- lm(Volume ~ Girth * Height,
               data = trees)

# perform BSE
step(lm_trees)
```

This BSE approach only gets as far as the first step (trying to drop the interaction term). We see immediately that dropping the interaction term makes the model worse and so the process stops. On the next line (underneath `Call:`) we see that the best model is still the full model and then we get to see the coefficients for each term.

### `airpoll` dataset

```{r, echo=FALSE}
airpoll <- read_csv("data/tidy/CS5-H2S.csv")
```

We construct a full linear model with both `treatment_plant`, `daily_temp` and the interaction between them and then we run the step function:

```{r}
# define the full model
lm_airpoll <- lm(hydrogen_sulfide ~ treatment_plant * daily_temp,
                 data = airpoll)

# perform BSE
step(lm_airpoll)
```

Again, this BSE approach only gets as far as the first step (trying to drop the interaction term). We see immediately that dropping the interaction term makes the model worse and so the process stops. On the next line (underneath `Call:`) we see that the best model is still the full model and then we get to see the coefficients for each term.

### `CO2` dataset

```{r}
# define the model, ignore the Plant variable
lm_co2 <- lm(uptake ~ Type + Treatment + conc
             + Type:Treatment + Type:conc + Treatment:conc
             + Type:Treatment:conc,
             data = CO2)
step(lm_co2)
```

This time we manage three steps. We first successful manage to drop the three-way interaction `Type:Treatment:conc`. At the next step we end up dropping the `Treatment:conc` interaction. At the final step we realise that we can't drop any more terms and so we're done. The minimal model here has 5 terms in it and the coefficients for this model are given at the very bottom of the output.

</details>
:::

## Key points

:::keypoints
- We can use Backwards Stepwise Elimination (BSE) on a full model to see if certain terms add to the predictive power of the model or not
- The AIC allows us to compare different models - if there is a difference in AIC of more than 2 between two models, then the smallest AIC score is more supported
- We can use the `step()` function to let R perform an automatic BSE
:::
