---
title: "Paired data"
---

```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python}
#| echo: false
#| message: false
import shutup;shutup.please()
exec(open('setup_files/setup.py').read())
```

::: {.callout-tip}
#### Learning outcomes

**Questions**

- When do I perform a paired two-sample test?
- What are the assumptions?
- How do I interpret and present the results of the test?
- How do I deal with paired non-normal data?

**Objectives**

- Set out your hypothesis for comparing two paired samples of continuous data
- Be able to summarise and visualise the data
- Understand and assess the underlying assumptions of the test
- Perform a paired two-sample t-test
- Be able to interpret and report the results
- Be able to do these steps on non-normal data

:::

A paired t-test is used when we have two samples of continuous data that can be paired (examples of these sort of data would be weights of individuals before and after a diet). This test is applicable if the number of paired points within the samples is large (\>30) or, if the number of points is small, then this test also works when the parent distributions are normally distributed.

## Libraries and functions

::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

### Libraries

```{r}
#| eval: false
# A collection of R packages designed for data science
library(tidyverse)

# Converts stats functions to a tidyverse-friendly format
library(rstatix)
```

### Functions

```{r}
#| eval: false
#| warning: false
# Performs a one-sample t-test, Student's t-test
# and Welch's t-test in later sections
rstatix::t_test()

# Performs a Shapiro-Wilk test for normality
stats::shapiro.test()

# Performs one and two sample Wilcoxon tests
rstatix::wilcox_test()

# Plots a Q-Q plot for comparison with a normal distribution
ggplot2::stat_qq()

# Adds a comparison line to the Q-Q plot
ggplot2::stat_qq_line()

# Plots jittered points by adding a small amount of random
# variation to each point, to handle overplotting
ggplot2::geom_jitter()

# Computes summary statistics                         
rstatix::get_summary_stats() 

# "Widens" the data, increasing the number of columns
tidyr::pivot_wider()
```

## Python

| Libraries  | Description                                                             |
|:----------------------|:------------------------------------------------|
| `plotnine` | The Python equivalent of `ggplot2`.                                     |
| `pandas`   | A Python data analysis and manipulation tool.                           |
| `pingouin` | A Python module developed to have simple yet exhaustive stats functions |

| Functions                                                                                     | Description                                                         |
|:-----------------------------------|:-----------------------------------|
| `pandas.DataFrame.read_csv`                                                                   | Reads in a `.csv` file                                              |
| `pandas.DataFrame.pivot()`                                                                    | Return reshaped DataFrame organised by given index / column values. |
| [`pingouin.normality()`](https://pingouin-stats.org/generated/pingouin.normality.html)        | Performs the Shapiro-Wilk test for normality.                       |
| [`pingouin.ttest()`](https://pingouin-stats.org/generated/pingouin.ttest.html#pingouin.ttest) | Performs a t-test                                                   |
| [`plotnine.stats.stat_qq()`](https://plotnine.readthedocs.io/en/stable/generated/plotnine.stats.stat_qq.html)           | Plots a Q-Q plot for comparison with a normal distribution.                                |
| [`plotnine.stats.stat_qq_line()`](https://plotnine.readthedocs.io/en/stable/generated/plotnine.stats.stat_qq_line.html) | Adds a comparison line to the Q-Q plot.                                                    |
:::
:::

## Data and hypotheses

For example, suppose we measure the cortisol levels in 20 adult females (nmol/l) first thing in the morning and again in the evening. We want to test whether the cortisol levels differs between the two measurement times. We will initially form the following null and alternative hypotheses:

-   $H_0$: There is no difference in cortisol level between times ($\mu M = \mu E$)
-   $H_1$: There is a difference in cortisol levels between times ($\mu M \neq \mu E$)

We use a two-sample, two-tailed paired t-test to see if we can reject the null hypothesis.

-   We use a **two-sample** test because we now have two samples
-   We use a **two-tailed** t-test because we want to know if our data suggest that the true (population) means are different from one another rather than that one mean is specifically bigger or smaller than the other
-   We use a **paired** test because each data point in the first sample can be linked to another data point in the second sample by a connecting factor
-   We're using a **t-test** because we're assuming that the parent populations are normal and have equal variance (We'll check this in a bit)

The data are stored in a tidy format in the file `data/CS1-twopaired.csv`.

::: {.panel-tabset group="language"}
## R

```{r}
# load the data
cortisol <- read_csv("data/CS1-twopaired.csv")

# have a look at the data
cortisol
```

## Python

```{python}
# load the data
cortisol_py = pd.read_csv('data/CS1-twopaired.csv')

# inspect the data
cortisol_py.head()
```
:::

We can see that the data frame consists of three columns:

1.  `patient_id`, a unique ID for each patient
2.  `time` when the cortisol level was measured
3.  `cortisol`, which contains the measured value.

For each `patient_id` there are two measurements: one in the morning and one in the afternoon.

## Summarise and visualise

It's always a good idea to visualise your data, so let's do that.

::: {.panel-tabset group="language"}
## R

```{r}
# create a boxplot
ggplot(cortisol,
       aes(x = time, y = cortisol)) +
  geom_boxplot() +
  geom_jitter(width = 0.05) +
  ylab("Cortisol level (nmol/l)")
```

Here we use also visualise the actual data points, to get a sense of how these data are spread out. To avoid overlapping the data points (try using `geom_point()` instead of `geom_jitter()`), we jitter the data points. What `geom_jitter()` does is add a small amount of variation to each point.

## Python

```{python}
#| results: hide
(ggplot(cortisol_py,
        aes(x = "time",
            y = "cortisol")) +
     geom_boxplot() +
     geom_jitter(width = 0.05) +
     ylab("Cortisol level (nmol/l)"))
```
:::

However, this plot does not capture how the cortisol level of each *individual* subject has changed though. We can explore the individual changes between morning and evening by looking at the *differences* between the two times of measurement for each patient.

To do this, we need to put our data into a wide format, so we can calculate the change in cortisol level for each patient.

::: {.panel-tabset group="language"}
## R

In `tidyverse` we can use the `pivot_wider()` function.

```{r}
# calculate the difference between evening and morning values
cortisol_diff <- cortisol %>%
  pivot_wider(id_cols = patient_id,
              names_from = time,
              values_from = cortisol) %>% 
  mutate(cortisol_change = evening - morning)

cortisol_diff
```

There are three arguments in `pivot_wider()`:

1. `id_cols = patient_id` tells it that each observational unit is determined by `patient_id`
2. `names_from = time` says that there will be new columns, with names from the `time` column (in this case, there are two values in there, `morning` and `evening`)
3. `values_from = cortisol` populates the new columns with the values coming from the `cortisol` column

Lastly, we create a new column `cortisol_change` that contains the difference between the `evening` and `morning` measurements.

After this we can plot our data:

```{r}
# plot the data
ggplot(cortisol_diff,
       aes(y = cortisol_change)) +
  geom_boxplot() +
  ylab("Change in cortisol (nmol/l)")
```

The differences in cortisol levels appear to be very much less than zero, meaning that the evening cortisol levels appear to be much lower than the morning ones. As such we would expect that the test would give a pretty significant result.

An alternative representation would be to plot the data points for both evening and morning and connect them by patient:

```{r}
# plot cortisol levels by patient
ggplot(cortisol,
       aes(x = time,
           y = cortisol,
           group = patient_id)) +
  geom_point() +
  geom_line()
```

This gives a similar picture to what the boxplot was telling us, that for most patients the cortisol levels are higher in the morning than in the evening.

## Python

```{python}
# reformat the data into a 'wide' format
cortisol_diff_py = pd.pivot(cortisol_py, index = "patient_id", columns = "time", values = "cortisol")

# add a new column with difference between
# evening and morning cortisol levels
cortisol_diff_py["cortisol_change"] = cortisol_diff_py["evening"].subtract(cortisol_diff_py["morning"])
      
# have a look at the format
cortisol_diff_py.head()
```

After this we can plot our data:

```{python}
#| results: hide
# plot the data
(ggplot(cortisol_diff_py,
        aes(x = "1",
            y = "cortisol_change")) +
     geom_boxplot() +
     ylab("Change in cortisol (nmol/l)"))
```

The differences in cortisol levels appear to be very much less than zero, meaning that the evening cortisol levels appear to be much lower than the morning ones. As such we would expect that the test would give a pretty significant result.

An alternative representation would be to plot the data points for both evening and morning and connect them by patient:

```{python}
#| results: hide
# plot cortisol levels by patient
(ggplot(cortisol_py,
        aes(x = "time",
            y = "cortisol",
            group = "patient_id")) +
     geom_point() +
     geom_line())
```

This gives a similar picture to what the boxplot was telling us, that for most patients the cortisol levels are higher in the morning than in the evening.
:::

## Assumptions

You will do this in the exercise!

## Implement and interpret the test

Perform a two-sample, two-tailed, paired t-test:

::: {.panel-tabset group="language"}
## R

```{r}
#| results: hide
# perform the test
t_test(cortisol ~ time,
       alternative = "two.sided",
       paired = TRUE,
       data = cortisol)
```

-   The first argument must be in the formula format: `response ~ predictor`
-   The `alternative` argument gives the type of alternative hypothesis and must be one of `two.sided`, `greater` or `less`
-   The `paired = TRUE` argument indicates that the data are paired

From our perspective the value of interest in the output is the `p` (`r t_test(cortisol ~ time, alternative = "two.sided", paired = TRUE, data = cortisol) |> pull(p)`).

## Python

To perform a paired t-test we can use the same `pg.ttest()` as before, but set the argument `paired = True`.

Annoyingly, the output is not entirely visible because the data frame is too wide. To deal with that, we can simply transpose it with `transpose()`

```{python}
pg.ttest(cortisol_diff_py["evening"],
         cortisol_diff_py["morning"],
         alternative = "two-sided",
         paired = True).transpose()
```

From our perspective the value of interest is the `p-val`.
:::

Since the p-value = 5.29 $\times$ 10<sup>-5</sup>) and thus substantially less than 0.05 we can reject the null hypothesis and state:

> A two-tailed, paired t-test indicated that the average cortisol level in adult females differed significantly between the morning (313.5 nmol/l) and the evening (197.4 nmol/l, p = 5.3 \* 10<sup>-5</sup>).

## Dealing with non-normal data

The example above assumes that the paired data come from parent distributions that are normal. As we've seen before, we may have data where we can't rely on that assumption. Fortunately, there is very little that we need to change in our approach if we want to analyse paired data that violate the assumption of normality.

As with the non-normal two-sample data, there is the underlying assumption that the parent distributions of the samples are comparable in shape and variance.

### Data and hypotheses

Using the `cortisol` data from before we form the following null and alternative hypotheses:

-   $H_0$: The median of the difference in cortisol levels between the two groups is 0 $(\mu M = \mu E)$
-   $H_1$: The median of the difference in cortisol levels between the two groups is not 0 $(\mu M \neq \mu E)$

We use a two-tailed Wilcoxon signed rank test to see if we can reject the null hypothesis.

### Summarise and visualise

Already implemented previously.

### Assumptions

These have been checked previously.

### Implement and interpret the test

Perform a two-tailed, Wilcoxon signed rank test:

::: {.panel-tabset group="language"}
## R

```{r}
wilcox_test(cortisol ~ time,
            alternative = "two.sided",
            paired = TRUE,
            data = cortisol)
```

-   The first argument must be in the formula format: `response ~ predictor`
-   The `alternative` argument gives the type of alternative hypothesis and must be one of `two.sided`, `greater` or `less`
- The `paired = TRUE` argument indicates that the data are paired

## Python

We'll use the wide format data set that we created previously:

```{python}
pg.wilcoxon(x = cortisol_diff_py["evening"],
            y = cortisol_diff_py["morning"],
            alternative = "two-sided",
            correction = True)
```
:::

The p-value is given in the `p` column (p-value = 0.000168). Given that this is less than 0.05 we can still reject the null hypothesis.

> A two-tailed, Wilcoxon signed rank test indicated that the median cortisol level in adult females differed significantly between the morning (320.5 nmol/l) and the evening (188.9 nmol/l, p = 0.00017).

## Exercises

### Cortisol levels {#sec-exr_cortisol}

:::{.callout-exercise}

{{< level 2 >}}

Check the assumptions necessary for this this paired t-test. Was a paired t-test an appropriate test?

::: {.callout-tip collapse="true"}
## Answer

We actually don't care too much about the distributions of the individual groups. Instead we care about the properties of the **differences**. So for a paired t-test to be valid for this data set, we need the differences between the morning and evening values to be normally distributed.

Let's check this with the Shapiro-Wilk test and Q-Q plots, using the wide data frames we created earlier.

::: {.panel-tabset group="language"}
## R

Perform Shapiro-Wilk test:

```{r}
# perform Shapiro-Wilk test on cortisol differences
shapiro.test(cortisol_diff$cortisol_change)
```

Create Q-Q plot:

```{r}
# create the Q-Q plot
ggplot(cortisol_diff,
       aes(sample = cortisol_change)) +
  stat_qq() +
  stat_qq_line(colour = "blue")
```

## Python

Perform Shapiro-Wilk test:

```{python}
# perform Shapiro-Wilk test on cortisol differences
pg.normality(cortisol_diff_py["cortisol_change"])
```

Create Q-Q plot:

```{python}
#| results: hide
# create the Q-Q plot
(ggplot(cortisol_diff_py,
        aes(sample = "cortisol_change")) +
     stat_qq() +
     stat_qq_line(colour = "red"))
```
:::

The Shapiro-Wilk test says that the data are normal enough and whilst the Q-Q plot is mostly fine, there is some suggestion of snaking at the bottom left. I'm actually OK with this because the suggestion of snaking is actually only due to a single point (the last point on the left). If you cover that point up with your thumb (or finger of your choice) then the remaining points in the Q-Q plot look pretty darn good, and so the suggestion of snaking is actually driven by only a single point (which can happen by chance). As such I'm happy that the assumption of normality is well-met in this case. This **single point** check is a useful thing to remember when assessing diagnostic plots.

So, yep, a paired t-test is appropriate for this data set.
:::
:::

### Deer legs {#sec-exr_deerlegs}

:::{.callout-exercise}

{{< level 2 >}}

Using the following data on deer legs (yes, really!), test the null hypothesis that the fore and hind legs of the deer in this data set are the same length.

```{r}
#| echo: false
#| warning: false
#| message: false
# deer leg length (cm)
read_csv("data/CS1-deer.csv") %>% 
  pivot_wider(names_from = leg, values_from = length) %>% 
  select(-id)
```

Do these results provide any evidence to suggest that fore- and hind-leg length differ in deer?

1.  Write down the null and alternative hypotheses
2.  Import the data from `data/CS1-deer.csv`
3.  Summarise and visualise the data
4.  Check your assumptions (normality and variance) using appropriate tests
5.  Discuss with your (virtual) neighbour which test is most appropriate?
6.  Perform the test
7.  Write down a sentence that summarises the results that you have found

::: {.callout-tip collapse="true"}
## Answer

#### Hypotheses

$H_0$ : foreleg average (mean or median) $=$ hindleg average (mean or median)

$H_1$ : foreleg average $\neq$ hindleg average

#### Import data, summarise and visualise

First of all, we need to load in the data.

::: {.panel-tabset group="language"}
## R

```{r}
#| message: false
#| warning: false
# load the data
deer <- read_csv("data/CS1-deer.csv")

# have a look
deer
```

The ordering of the data is important here; the first hind leg row corresponds to the first fore leg row, the second to the second and so on. To indicate this we use an `id` column, where each observation has a unique ID.

Let's look at the data and see what it tells us.

```{r}
# summarise the data
summary(deer)
```

We can also summarise some of the main summary statistics for each type of `leg`. We don't need summary statistics for the `id` column, so we unselect it with `select(-id)`.

To make life easy we use the `get_summary_stats()` function from the `rstatix` package. Have a look at the help function to see what kind of summary statistics it can produce. In this case I'm using the `type = "common"` option to specify that I want to find commonly used statistics (e.g. sample number, min, max, median, mean etc.)

```{r}
# or even summarise by leg type
deer %>% 
  select(-id) %>% 
  group_by(leg) %>% 
  get_summary_stats(type = "common")
```

Visualising the data is often more useful:

```{r}
# we can also visualise the data
ggplot(deer,
       aes(x = leg, y = length)) +
    geom_boxplot()
```

All of this suggests that there might be a difference between the legs, with hind legs being longer than forelegs. However, this representation obscures the fact that we have *paired* data. What we really need to look at is the _difference_ in leg length for each observation:

```{r}
# create a data set that contains the difference in leg length
leg_diff <- deer %>% 
  pivot_wider(id_cols = id,
              names_from = leg,
              values_from = length) %>% 
  mutate(leg_diff = hindleg - foreleg)
```

```{r}
# plot the difference in leg length
ggplot(leg_diff,
       aes(y = leg_diff)) +
  geom_boxplot()
```

Additionally, we can also plot the data by observation:

```{r}
# plot the data by observation
ggplot(deer,
       aes(x = leg, y = length, group = id)) +
  geom_point() +
  geom_line()
```

## Python

```{python}
#| message: false
#| warning: false
# load the data
deer_py = pd.read_csv("data/CS1-deer.csv")

# have a look
deer_py.head()
```

The ordering of the data is important here; the first hind leg row corresponds to the first fore leg row, the second to the second and so on. To indicate this we use an `id` column, where each observation has a unique ID.

Let's look at the data and see what we can see.

```{python}
# summarise the data
deer_py.describe()
```

We can also summarise by leg type:

```{python}
deer_py.groupby("leg")["length"].describe()
```

It might be more helpful to look at the *difference* in leg length. In order to calculate that, we need to reformat our data into a 'wide' format first:

```{python}
# reformat the data into a 'wide' format
leg_diff_py = pd.pivot(deer_py,
                       index = "id",
                       columns = "leg",
                       values = "length")

# have a look at the format
leg_diff_py.head()
```

Next, we can add a new column `leg_diff` that contains the leg difference:

```{python}
# add a new column with difference between
# hind and fore leg length
leg_diff_py["leg_diff"] = leg_diff_py["hindleg"].subtract(leg_diff_py["foreleg"])
 
```

Finally, we can visualise this:

```{python}
#| results: hide
# we can also visualise the data
(ggplot(leg_diff_py,
        aes(x = "1",
            y = "leg_diff")) +
     geom_boxplot())
```

All of this suggests that there might be a difference between the legs, with hind legs being longer than forelegs. However, this representation obscures the fact that we have *paired* data. What we really need to look at is the difference in leg length for each observation:

```{python}
#| results: hide
# plot paired observations
(ggplot(deer_py,
        aes(x = "leg",
            y = "length",
            group = "id")) +
     geom_point() +
     geom_line())
```
:::

All of this gives us a much clearer picture. It looks as though the hindlegs are about 4 cm longer than the forelegs, on average. It also suggests that our leg differences might not be normally distributed (the data look a bit skewed in the boxplot).

#### Assumptions

We need to consider the distribution of the *difference* in leg lengths rather than the individual distributions.

::: {.panel-tabset group="language"}
## R

Shapiro-Wilk test:

```{r}
# perform Shapiro-Wilk test on leg differences
shapiro.test(leg_diff$leg_diff)
```

Q-Q plot:

```{r}
# create a Q-Q plot
ggplot(leg_diff,
       aes(sample = leg_diff)) +
  stat_qq() +
  stat_qq_line(colour = "blue")
```

## Python

Shapiro-Wilk test:

```{python}
# perform Shapiro-Wilk test on leg length differences
pg.normality(leg_diff_py["leg_diff"])
```

Create the Q-Q plot:

```{python}
#| results: hide
# create the Q-Q plot
(ggplot(leg_diff_py,
        aes(sample = "leg_diff")) +
     stat_qq() +
     stat_qq_line(colour = "red"))
```
:::

Both our Shapiro-Wilk test and our Q-Q plot suggest that the difference data aren't normally distributed, which rules out a paired t-test. We should therefore consider a paired Wilcoxon signed rank test next. Remember that this test requires that the distribution of differences be of a similar shape, whereas our box plot from before suggested that the data were very much skewed.

This means that we're not able to perform a paired Wilcoxon signed rank test either!

#### Conclusions

So, frustratingly, neither of the tests at our disposal are appropriate for this data set. The differences in fore leg and hind leg lengths are neither normal enough for a paired t-test nor are they symmetric enough for a Wilcoxon signed rank test. We also don't have enough data to just use the t-test (we'd need more than 30 points or so). So what do we do in this situation? Well, the answer is that there aren't actually any traditional statistical tests that are valid for this data set as it stands!

There are two options available to someone:

1.  try transforming the raw data (take logs, square root, reciprocals) and hope that one of them leads to a modified data set that satisfies the assumptions of one of the tests we've covered, or
2.  use a permutation test approach (which would work but is beyond the scope of this course).

The reason I included this example in the first practical is purely to illustrate how a very simple data set with an apparently clear message (leg lengths differ within deer) can be intractable. You don't need to have very complex data sets before you go beyond the capabilities of classical statistics.

As Jeremy Clarkson [would put it](https://www.quotes.net/mquote/941330):

> And on that bombshell, it's time to end. Goodnight!

:::
:::

## Summary

::: {.callout-tip}
#### Key points

-   Paired t-tests are used when you have two paired samples of continuous data, which are normally distributed and have equal variance
-   A good way of assessing the assumption of normality is by checking the data against a Q-Q plot
- We can check equality of variance (homoscedasticity) with Bartlett's (normal data) or Levene's (non-normal data) test
-   The Wilcoxon signed rank test is used when you have two paired samples of continuous data, which are not normally distributed (but have comparable distributional shapes), and have equal variance.
:::

