---
title: "ANOVA"
---

```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python}
#| echo: false
#| message: false
import shutup;shutup.please()
exec(open('setup_files/setup.py').read())
```

::: {.callout-tip}
#### Learning outcomes

**Questions**

-   How do I analyse multiple samples of continuous data?
-   What is an ANOVA?
-   How do I check for differences between groups?

**Objectives**

-   Be able to perform an ANOVA in R
-   Understand the ANOVA output and evaluate the assumptions
-   Understand what post-hoc testing is and how to do this in R
:::

## Purpose and aim

Analysis of variance or ANOVA is a test than can be used when we have multiple samples of continuous response data. Whilst it is possible to use ANOVA with only two samples, it is generally used when we have three or more groups. It is used to find out if the samples came from parent distributions with the same mean. It can be thought of as a generalisation of the two-sample Student's t-test.

## Libraries and functions

::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

### Libraries

```{r}
#| eval: false
# A collection of R packages designed for data science
library(tidyverse)

# Converts stats functions to a tidyverse-friendly format
library(rstatix)

# Creates diagnostic plots using ggplot2
library(ggResidpanel)
```

### Functions

```{r}
#| eval: false
#| warning: false
# Computes summary statistics
rstatix::get_summary_stats()

# Perform Tukey's range test
rstatix::tukey_hsd()

# Creates diagnostic plots
ggResidpanel::resid_panel()

# Fits a linear model  
stats::lm()

# Carries out an ANOVA on a linear model 
stats::anova()

# Performs a Shapiro-Wilk test for normality
stats::shapiro.test()
```

## Python

### Libraries

```{python}
#| eval: false
# A Python data analysis and manipulation tool
import pandas as pd

# Simple yet exhaustive stats functions.
import pingouin as pg

# Python equivalent of `ggplot2`
from plotnine import *

# Statistical models, conducting tests and statistical data exploration
import statsmodels.api as sm

# Convenience interface for specifying models using formula strings and DataFrames
import statsmodels.formula.api as smf
```

### Functions

```{python}
#| eval: false
# Summary statistics
pandas.DataFrame.describe()

# Plots the first few rows of a DataFrame
pandas.DataFrame.head()

# Query the columns of a DataFrame with a boolean expression
pandas.DataFrame.query()

# Reads in a .csv file
pandas.read_csv

# Performs an analysis of variance
pingouin.anova()

# Tests for equality of variance
pingouin.homoscedasticity()

# Performs the Shapiro-Wilk test for normality
pingouin.normality()

# Creates a model from a formula and data frame
statsmodels.formula.api.ols

# Creates an ANOVA table for one or more fitted linear models
statsmodels.stats.anova.anova_lm

# Custom function to create diagnostic plots
dgplots()
```

Note: you can download the `dgplots` script [here](scripts/dgplots.py).

:::
:::

## Data and hypotheses

For example, suppose we measure the feeding rate of oyster catchers (shellfish per hour) at three sites characterised by their degree of shelter from the wind, imaginatively called `exposed` (E), `partially sheltered` (P) and `sheltered` (S). We want to test whether the data support the hypothesis that feeding rates don't differ between locations. We form the following null and alternative hypotheses:

-   $H_0$: The mean feeding rates at all three sites is the same $\mu E = \mu P = \mu S$
-   $H_1$: The mean feeding rates are not all equal.

We will use a one-way ANOVA test to check this.

-   We use a **one-way** ANOVA test because we only have one predictor variable (the categorical variable location).
-   We're using **ANOVA** because we have more than two groups and we don't know any better yet with respect to the exact assumptions.

The data are stored in the file `data/CS2-oystercatcher-feeding.csv`.

## Summarise and visualise

::: {.panel-tabset group="language"}
## R

First we read in the data.

```{r}
#| message: false
#| warning: false
# load data
oystercatcher <- read_csv("data/CS2-oystercatcher-feeding.csv")

# and have a look
oystercatcher
```

The `oystercatcher` data set contains two columns:

1.  a `site` column with information on the amount of shelter of the feeding location
2.  a `feeding` column containing feeding rates

Next, we get some basic descriptive statistics:

```{r}
# get some basic descriptive statistics
oystercatcher %>% 
  group_by(site) %>% 
  get_summary_stats(type = "common")
```

Finally, we plot the data by `site`:

```{r}
# plot the data
ggplot(oystercatcher,
       aes(x = site, y = feeding)) +
  geom_boxplot()
```

## Python

First, we read in the data.

```{python}
# load the data
oystercatcher_py = pd.read_csv("data/CS2-oystercatcher-feeding.csv")

# and have a look
oystercatcher_py.head()
```

The `oystercatcher_py` data set contains two columns:

1.  a `site` column with information on the amount of shelter of the feeding location
2.  a `feeding` column containing feeding rates

Next, we get some basic descriptive statistics *per group*. Here we use the `pd.groupby()` function to group by `site`. We only want to have summary statistics for the `feeding` variable, so we specify that as well:

```{python}
oystercatcher_py.groupby("site")["feeding"].describe()
```

Finally, we plot the data:

```{python}
#| results: hide
# plot the data
p = (ggplot(oystercatcher_py,
        aes(x = "site",
            y = "feeding")) +
     geom_boxplot())

p.show()
```
:::

Looking at the data, there appears to be a noticeable difference in feeding rates between the three sites. We would probably expect a reasonably significant statistical result here.

## Assumptions

To use an ANOVA test, we have to make three assumptions:

1.  The parent distributions from which the samples are taken are normally distributed
2.  Each data point in the samples is independent of the others
3.  The parent distributions should have the same variance

In a similar way to the two-sample tests we will consider the normality and equality of variance assumptions both using tests and by graphical inspection (and ignore the independence assumption).

### Normality

First we perform a Shapiro-Wilk test on each site separately.

::: {.panel-tabset group="language"}
## R

We take the data, filter for each type of `site`, extract the `feeding` rates and send those data to the `shapiro.test()` function.

```{r}
# Shapiro-Wilk test on each site
oystercatcher %>% 
    filter(site == "exposed") %>% 
    pull(feeding) %>% 
    shapiro.test()

oystercatcher %>% 
    filter(site == "partial") %>% 
    pull(feeding) %>% 
    shapiro.test()

oystercatcher %>% 
    filter(site == "sheltered") %>% 
    pull(feeding) %>% 
    shapiro.test()
```

## Python

We use the `pg.normality()` function to calculate the statistic. This requires:

-   the `dv` dependent variable (`feeding` in our case)
-   the `group` variable (`site`)
-   and some data

```{python}
pg.normality(dv = "feeding",
             group = "site",
             data = oystercatcher_py)
```
:::

We can see that all three groups appear to be normally distributed which is good.

::: {.callout-important}
For ANOVA however, considering each group in turn is often considered quite excessive and, in most cases, it is sufficient to consider the normality of the combined **set of residuals** from the data. We'll explain residuals properly in the [next session](#cs3-intro), but effectively they are the difference between each data point and its group mean.

To get hold of these residuals, we need to create a **linear model**. Again, this will be explained in more detail in the next section. For now, we see it as a way to describe the relationship between the `feeding` rate and the `site`.
:::

So, we create a linear model, extract the residuals and check their normality:

::: {.panel-tabset group="language"}
## R

We use the `lm()` function to define the linear model that describes the relationship between `feeding` and `site`. The notation is similar to what we used previously when we were dealing with two samples of data.

```{r}
# define the model
lm_oystercatcher <- lm(feeding ~ site,
                       data = oystercatcher)
```

We can read this as "create a linear model (`lm`) where the feeding rate (`feeding`) depends on the site (`site`), using the `oystercatcher` data".

We store the output of that in an object called `lm_oystercatcher`. We'll look into what this object contains in more detail in later sections.

For now, we extract the residuals from this object using the `residuals()` function and then use this in the `shapiro.test()` function.

``` {r}
# extract the residuals
resid_oyster <- residuals(lm_oystercatcher)

# perform Shapiro-Wilk test on residuals
shapiro.test(resid_oyster)
```

## Python

Unfortunately `pingouin` does not have a straightforward way of extracting residuals (if you know more, please let me know!).

To get our residuals we use `statsmodels`, a module that provides functions for statistical models. We'll be using this in upcoming sessions, so you'll have a head start!

At this point you shouldn't concern yourself too much with the exact syntax, just run it an have a look.

::: {.callout-note collapse="true"}
## Technical details (optional)

We need to import a few extra modules. First, we load the `statsmodels.api` module, which contains an `OLS()` function (Ordinary Least Squares - the equivalent of the `lm()` function in R).

We also import `stats.models.formula.api` so we can use the formula notation in our linear model. We define the formula as `formula = "feeding ~ C(site)"` with `C` conveying that the `site` variable is a category. Lastly we can `.fit()` the model.

If you're familiar with this stuff then you can look at the model itself by running `summary(lm_oystercatcher_py)`. But we'll cover all of this in later sessions.
:::

We load the modules, define a linear model, create a `fit()` and we get the residuals from the linear model fit with `.resid`.

```{python}
#| eval: false
import statsmodels.api as sm
import statsmodels.formula.api as smf
```

```{python}
# create a linear model
model = smf.ols(formula = "feeding ~ C(site)", data = oystercatcher_py)
# and get the fitted parameters of the model
lm_oystercatcher_py = model.fit()
```

```{python}
# get the residuals from the model fit
# and perform Shapiro-Wilk test
pg.normality(lm_oystercatcher_py.resid)
```
:::

Again, we can see that the combined residuals from all three groups appear to be normally distributed (which is as we would have expected given that they were all normally distributed individually!)

### Equality of variance

We now test for equality of variance using Bartlett's test (since we've just found that all of the individual groups are normally distributed).

Perform Bartlett's test on the data:

::: {.panel-tabset group="language"}
## R

```{r}
# check equality of variance
bartlett.test(feeding ~ site,
              data = oystercatcher)
```

Where the relevant p-value is given on the 3rd line. Here we see that each group appears to have comparable variance.

## Python

We use the `homoscedasticity()` function from `pingouin` (homoscedasticity is another way of describing equality of variance). The default `method` is `levene`, so we need to specify that we want to use `bartlett`.

```{python}
pg.homoscedasticity(dv = "feeding",
                    group = "site",
                    method = "bartlett",
                    data = oystercatcher_py)
```

Where the relevant p-value is given in the `pval` column. Here we see that each group appears to have the same variance.
:::

### Graphical interpretation and diagnostic plots

Assessing assumptions via these tests can be cumbersome, but also a bit misleading at times. It reduces the answer to the question "is the assumption met?" to a yes/no, based on some statistic and associated p-value.

This does not convey that things aren't always so clear-cut and that there is a lot of grey area that we need to navigate. As such, assessing assumptions through graphical means - using diagnostic plots - is often preferred.

::: {.panel-tabset group="language"}
## R

In the first session we already created diagnostic Q-Q plots directly from our data, using `stat_qq()` and `stat_qq_line()`. For more specific plots this becomes a bit cumbersome. There is an option to create ggplot-friendly diagnostic plots, using the `ggResidPanel` package.

If you haven't got `ggResidpanel` installed, please run the following code:

```{r}
#| eval: false
# install package
install.packages("ggResidpanel")

# load library
library(ggResidpanel)
```

Let's create the diagnostic plots we're interested in using `ggResidPanel`. It takes a linear model object as input (`lm_oystercatcher`) and you can define which plots you want to see using the `plots = ` argument. I have also added a smoother line (`smoother = TRUE`) to the plots, which we'll use to compare against.

```{r}
#| message: false
resid_panel(lm_oystercatcher,
            plots = c("resid", "qq", "ls", "cookd"),
            smoother = TRUE)
```

-   The top left graph plots the **Residuals plot**. If the data are best explained by a linear line then the points should be uniformly distributed above and below the horizontal blue line. If that's the case then the red line (a smoother line) should overlay the blue line. This plot looks pretty good.
-   The top right graph shows the **Q-Q plot** which allows a visual inspection of normality. If the residuals are normally distributed, then the points should lie on the diagonal blue line. This plot looks good.
-   The bottom left **Location-Scale** graph allows us to investigate whether there is any correlation between the residuals and the predicted values and whether the variance of the residuals changes significantly. If not, then the red line should be horizontal. If there is any correlation or change in variance then the red line will not be horizontal. This plot is fine.
-   The last graph shows the **Cook's distance** and tests if any one point has an unnecessarily large effect on the fit. A rule of thumb is that if any value is larger than 1.0, then it might have a large effect on the model. If not, then no point has undue influence. This plot is good. There are different ways to determine the threshold (apart from simply setting it to 1) and in this plot the blue dashed line is at `4/n`, with `n` being the number of samples. At this threshold there are some data points that may be influential, but I personally find this threshold rather strict.

## Python

Unfortunately Python doesn't provide a convenient way of displaying the same diagnostic plots as R does.

I created a function `dgplots()` (which stands for Diagnostic Plots - very original I know...) that does this for you. All you need to do is create a linear model, get the fit and feed that to the `dgplots()` function.

You can find the script for this [here](scripts/dgplots.py).

You can of course plot the model values yourself by extracting them from the linear model fit, but this should provide a convenient way to avoid that kind of stuff.

```{python}
dgplots(lm_oystercatcher_py)
```

-   The top left graph plots the **Residuals plot**. If the data are best explained by a linear line then the points should be uniformly distributed above and below the horizontal blue line. If that's the case then the red line (a smoother line) should overlay the blue line. This plot looks pretty good.
-   The top right graph shows the **Q-Q plot** which allows a visual inspection of normality. If the residuals are normally distributed, then the points should lie on the diagonal blue line. This plot looks good.
-   The bottom left **Location-Scale** graph allows us to investigate whether there is any correlation between the residuals and the predicted values and whether the variance of the residuals changes significantly. If not, then the red line should be horizontal. If there is any correlation or change in variance then the red line will not be horizontal. This plot is fine.
-   The last graph shows the **Influential points** and tests if any one point has an unnecessarily large effect on the fit. Here we're using the Cook's distance as a measure. A rule of thumb is that if any value is larger than 1.0, then it might have a large effect on the model. If not, then no point has undue influence. This plot is good. There are different ways to determine the threshold (apart from simply setting it to 1) and in this plot the blue dashed line is at `4/n`, with `n` being the number of samples. At this threshold there are some data points that may be influential, but I personally find this threshold rather strict.
:::

We can see that these graphs are very much in line with what we've just looked at using the test, which is reassuring. The groups all appear to have the same spread of data, and the Q-Q plot shows that the assumption of normality is alright.

::: callout-important
## Assessing assumptions

At this stage, I should point out that I nearly always stick with the graphical method for assessing the assumptions of a test. Assumptions are rarely either completely met or not met and there is always some degree of personal assessment.

Whilst the formal statistical tests (like Shapiro-Wilk) are technically fine, they can often create a false sense of things being absolutely right or wrong in spite of the fact that they themselves are still probabilistic statistical tests. In these exercises we are using both approaches whilst you gain confidence and experience in interpreting the graphical output and whilst it is absolutely fine to use both in the future I would strongly recommend that you don't rely solely on the statistical tests in isolation.
:::

## Implement and interpret the test

As is often the case, performing the actual statistical test is the least of your efforts. 

::: {.panel-tabset group="language"}
## R

In R we perform the ANOVA on the linear model object, `lm_oystercatcher` in this case. We do this with the `anova()` function:

```{r}
anova(lm_oystercatcher)
```

This takes the linear model (i.e. finds the means of the three groups and calculates a load of intermediary data that we need for the statistical analysis) and stores this information in an R object (which we've called `lm_oystercatcher`, but which you can call what you like).

In the output:

-   The 1st line just tells you the that this is an ANOVA test
-   The 2nd line tells you what the response variable is (in this case feeding)
-   The 3rd, 4th and 5th lines are an ANOVA table which contain some useful values:
    -   The `Df` column contains the degrees of freedom values on each row, `r glance(lm_oystercatcher) %>% pull(df) %>% as.numeric()` and `r glance(lm_oystercatcher) %>% pull(df.residual) %>% as.numeric()` (which we can use for the reporting)
    -   The `F` value column contains the F statistic, `r round(glance(lm_oystercatcher) %>% pull(statistic) %>% as.numeric(), digits = 2)` (which again we'll need for reporting).
    -   The p-value is 2.2e-16 and is the number directly under the `Pr(>F)` on the 4th line (to be precise, it is `r formatC((glance(lm_oystercatcher) %>% pull(p.value) %>% as.numeric()), format = "e", digits = 2)` but anything smaller than 2.2e-16 gets reported as `< 2.2e-16`).
    -   The other values in the table (in the `Sum Sq` and `Mean Sq`) columns are used to calculate the F statistic itself and we don't need to know these.
-   The 6th line has some symbolic codes to represent how big (small) the p-value is; so, a p-value smaller than 0.001 would have a \*\*\* symbol next to it (which ours does). Whereas if the p-value was between 0.01 and 0.05 then there would simply be a \* character next to it, etc. Thankfully we can all cope with actual numbers and don't need a short-hand code to determine the reporting of our experiments (please tell me that's true...!)

## Python

There are different ways of conducting an ANOVA in Python, with `scipy.stats` [proving an option](https://docs.scipy.org/doc/scipy/reference/generated/scipy.stats.f_oneway.html). However, I find that the [anova()](https://pingouin-stats.org/generated/pingouin.anova.html#pingouin.anova) function in `pingouin` provides the easiest and most-detailed option to do this.

It takes the following arguments:

-   `dv`: dependent variable (response variable; in our case `feeding`)
-   `between`: between-subject factor (predictor variable; in our case `site`)
-   `data`: which function doesn't!?
-   `detailed`: optional `True` or `False`, we're setting it to `True` because we like to know what we're doing!

```{python}
pg.anova(dv = "feeding",
         between = "site",
         data = oystercatcher_py,
         detailed = True)
```

This creates a linear model based on the data, *i.e*. finds the means of the three groups and calculates a load of intermediary data that we need for the statistical analysis.

In the output:

-   `Source`: Factor names - in our case these are the different sites (`site`)
-   `SS`: Sums of squares (we'll get to that in a bit)
-   `DF`: Degrees of freedom (at the moment only used for reporting)
-   `MS`: Mean squares
-   `F`: Our F-statistic
-   `p-unc`: p-value (`unc` stands for "uncorrected" - more on multiple testing correction later)
-   `np2`: Partial eta-square effect sizes (more on this later)

Alternatively, and we'll be using this method later on in the course, you can perform an ANOVA on the `lm_oystercatcher_py` object we created earlier.

This uses the `sm.stats.anova_lm()` function from `statsmodels`. As you'll see, the output is very similar:

```{python}
sm.stats.anova_lm(lm_oystercatcher_py)
```
:::

Again, the p-value is what we're most interested in here and shows us the probability of getting samples such as ours if the null hypothesis were actually true.

Since the p-value is very small (much smaller than the standard significance level of 0.05) we can say "that it is very unlikely that these three samples came from the same parent distribution" and as such we can reject our null hypothesis and state that:

> A one-way ANOVA showed that the mean feeding rate of oystercatchers differed significantly between locations (p = `r formatC((glance(lm_oystercatcher) %>% pull(p.value) %>% as.numeric()), format = "e", digits = 2)`).

## Post-hoc testing (Tukey's range test)

One drawback with using an ANOVA test is that it only tests to see if all of the means are the same. If we get a significant result using ANOVA then all we can say is that not all of the means are the same, rather than anything about how the pairs of groups differ. For example, consider the following box plot for three samples.

```{r}
#| echo: false
#| message: false
# load the data
tukey <- read_csv("data/CS2-tukey.csv")
```

```{r}
#| echo: false
ggplot(tukey,
       aes(x = group, y = response)) +
    geom_boxplot()
```

Each group is a random sample of 20 points from a normal distribution with variance 1. Groups 1 and 2 come from a parent population with mean 0 whereas group 3 come from a parent population with mean 2. The data clearly satisfy the assumptions of an ANOVA test.

How do we find out if there are any differences between these groups and, if so, which groups are different from each other?

### Read in data and plot

::: {.panel-tabset group="language"}
## R

```{r}
# load the data
tukey <- read_csv("data/CS2-tukey.csv")

# have a look at the data
tukey
```

```{r}
# plot the data
ggplot(tukey,
       aes(x = group, y = response)) +
    geom_boxplot()
```

## Python
```{python}
# load the data
tukey_py = pd.read_csv("data/CS2-tukey.csv")

# have a look at the data
tukey_py.head()
```

```{python}
#| results: hide
# plot the data
p = (ggplot(tukey_py,
        aes(x = "group",
            y = "response")) +
     geom_boxplot())

p.show()
```

:::

### Test for a significant difference in group means

::: {.panel-tabset group="language"}
## R

```{r}
# create a linear model
lm_tukey <- lm(response ~ group,
               data = tukey)

# perform an ANOVA
anova(lm_tukey)
```

## Python

```{python}
pg.anova(dv = "response",
         between = "group",
         data = tukey_py,
         detailed = True)
```
:::

Here we have a p-value of 2.39 $\times$ 10<sup>-7</sup> and so the test has very conclusively rejected the hypothesis that all means are equal.

However, this was not due to all of the sample means being different, but rather just because one of the groups is very different from the others. In order to drill down and investigate this further we use a new test called **Tukey's range test** (or **Tukey's honest significant difference test** -- this always makes me think of some terrible cowboy/western dialogue).

This will compare all of the groups in a pairwise fashion and reports on whether a significant difference exists.

### Performing Tukey's test

::: {.panel-tabset group="language"}
## R

To perform Tukey's range test we can use the `tukey_hsd()` function from the `rstatix` package. Note, there is a `TukeyHSD()` function in base R as well, but the `tukey_hsd()` function can take a linear model object as input, whereas the `TukeyHSD()` function cannot. This makes the `tukey_hsd()` function a bit easier to work with.

```{r}
# perform Tukey's range test on linear model
tukey_hsd(lm_tukey)
```

The `tukey_hsd()` function takes our linear model (`lm_tukey`) as its input. The output is a pair-by-pair comparison between the different groups (samples 1 to 3). We are interested in the `p.adj` column, which gives us the adjusted p-value. The null hypothesis in each case is that there is no difference in the mean between the two groups.

## Python

```{python}
pg.pairwise_tukey(dv = "response",
                  between = "group",
                  data = tukey_py).transpose()
```

The `dv` argument is the response variable, whereas the `between` argument defines the explanatory variable.

We `.transpose()` the data, so we can look at the output a bit easier. Doing so, we focus on the `p-tukey` values.
:::

As we can see that there isn't a significant difference between `sample1` and `sample2` but there is a significant difference between `sample1` and `sample3`, as well as `sample2` and `sample3`. This matches with what we expected based on the box plot.

### Assumptions

When to use Tukey's range test is a matter of debate (strangely enough a lot of statistical analysis techniques are currently matters of opinion rather than mathematical fact -- it does explain a little why this whole field appears so bloody confusing!)

-   Some people claim that we should only perform Tukey's range test (or any other post-hoc tests) if the preceding ANOVA test showed that there was a significant difference between the groups and that if the ANOVA test had not shown any significant differences between groups then we would have to stop there.
-   Other people say that this is rubbish and we can do whatever we like as long as we tell people what we did.

The background to this is rather involved but one of the reasons for this debate is to prevent so-called **data-dredging** or **p-hacking**. This is where scientists/analysts are so fixated on getting a "significant" result that they perform a huge variety of statistical techniques until they find one that shows that their data is significant (this was a particular problem in psychological studies for while -- not to point fingers though, they are working hard to sort their stuff out. Kudos!).

Whether you should use post-hoc testing or not will depend on your experimental design and the questions that you're attempting to answer.

Tukey's range test, when we decide to use it, requires the same three assumptions as an ANOVA test:

1.  Normality of distributions
2.  Equality of variance between groups
3.  Independence of observations

## Exercises

::: {.callout-exercise}
#### Lobster weight {#sec-exr_lobsters}
{{< level 2 >}}

Juvenile lobsters in aquaculture were grown on three different diets (fresh mussels, semi-dry pellets and dry flakes). After nine weeks, their wet weight was:

```{r}
#| echo: false
#| message: false
#| warning: false
read_csv("data/examples/cs2-messy_lobsters.csv") %>%
    pivot_wider(names_from = diet, values_from = weight) %>%
    select(-group)
```

Is there any evidence that diet affects the growth rate of lobsters?

1.  Write down the null and alternative hypotheses
2.  Import the data from `data/CS2-lobsters.csv`
3.  Summarise and visualise the data
4.  Check the assumptions using appropriate tests and graphical analyses
5.  Perform an ANOVA
6.  Write down a sentence that summarises the results that you have found
7.  Perform a post-hoc test and report the findings

::: {.callout-answer}

#### Hypotheses

$H_0$ : all means are equal

$H_1$ : not all means are equal

#### Import Data, summarise and visualise

::: {.panel-tabset group="language"}
## R

```{r}
#| message: false
#| warning: false
# load the data
lobsters <- read_csv("data/CS2-lobsters.csv")

# look at the data
lobsters
```

The data have a unique `id` column, which we don't need any summary statistics for, so we deselect it:

```{r}
# create some summary statistics
lobsters %>% 
  select(-id) %>% 
  group_by(diet) %>% 
  get_summary_stats(type = "common")
```

Next, we visualise the data:

```{r}
lobsters %>% 
  ggplot(aes(x = diet, y = weight)) +
  geom_boxplot()
```

## Python

```{python}
lobsters_py = pd.read_csv("data/CS2-lobsters.csv")
```

Next, we visualise the data:

```{python}
#| results: hide
p = (ggplot(lobsters_py,
        aes(x = "diet",
            y = "weight")) +
     geom_boxplot())

p.show()
```

:::

As always we use the plot and summary to assess three things:

1.  Did we load the data in properly?

-   We see three groups with reasonable values. There aren't any data points that are obviously wrong (negative, zero or massively big) and we have the right number of groups. So it looks as if we didn't do anything obviously wrong.

2.  What do we expect as a result of a statistical test?

-   Whilst the `Mussels` group does look higher than the other two groups, `Pellets` and `Flakes` appear almost identical in terms of average values, and there's quite a bit of overlap with the `Mussels` group. A non-significant result is the most likely answer, and I would be surprised to see a significant p-value - especially given the small sample size that we have here.

3.  What do we think about assumptions?

-   The groups appear mainly symmetric (although `Pellets` is a bit weird) and so we're not immediately massively worried about lack of normality. Again, `Flakes` and `Mussels` appear to have very similar variances but it's a bit hard to decide what's going on with `Pellets.` It's hard to say what's going on with the assumptions and so I'll wait and see what the other tests say.

#### Explore Assumptions

We'll explore the assumption of normality and equality of variance, assuming that the data are independent.

::: {.panel-tabset group="language"}
## R

**Normality**

We'll be really thorough here and consider the normality of each group separately and jointly using the Shapiro-Wilk test, as well as looking at the Q-Q plot. In reality, and after these examples , we'll only use the Q-Q plot.

First, we perform the Shapiro-Wilk test on the individual groups:

```{r}
# Shapiro-Wilk on lobster groups
lobsters %>% 
    filter(diet == "Flakes") %>% 
    pull(weight) %>% 
    shapiro.test()

lobsters %>% 
    filter(diet == "Mussels") %>% 
    pull(weight) %>% 
    shapiro.test()

lobsters %>% 
    filter(diet == "Pellets") %>% 
    pull(weight) %>% 
    shapiro.test()
```

`Flakes` and `Mussels` are fine, but, as we suspected from earlier, `Pellets` appears to have a marginally significant Normality test result.

Let's look at the Shapiro-Wilk test for all of the data together:

```{r}
# create a linear model
lm_lobsters <- lm(weight ~ diet,
                  data = lobsters)

# extract the residuals
resid_lobsters <- residuals(lm_lobsters)

# and perform the Shapiro-Wilk test on the residuals
resid_lobsters %>% 
  shapiro.test()
```

This on the other hand says that everything is fine. Let's look at Q-Q plot.

```{r}
# Q-Q plots
resid_panel(lm_lobsters,
            plots = "qq")
```

Here, I've used an extra argument to the normal diagnostic plots call. The default option is to plot 4 diagnostic plots. You can tell `resid_panel()` to only plot a specific one, using the `plots =` arguments. If you want to know more about this have a look at the [help documentation](https://goodekat.github.io/ggResidpanel-tutorial/tutorial.html#overview) or by using `?resid_panel`.

The Q-Q plot looks OK, not perfect, but more than good enough for us to have confidence in the normality of the data.

Overall, I'd be happy that the assumption of normality has been adequately well met here. The suggested lack of normality in the `Pellets` was only just significant and we have to take into account that there are only 5 data points in that group. If there had been a lot more points in that group, or if the Q-Q plot was considerably worse then I wouldn't be confident.

**Equality of Variance**

We'll consider the Bartlett test and we'll look at some diagnostic plots too.

```{r}
#| message: false
# perform Bartlett's test
bartlett.test(weight ~ diet,
              data = lobsters)

# plot the residuals and scale-location plots
resid_panel(lm_lobsters,
            plots = c("resid", "ls"),
            smoother = TRUE)
```

In the above code I've specified which diagnostic plots I wanted. I have also added a smoother line (`smoother = TRUE`) to the plots.

1.  The Residuals Plot. What we're looking for there is that the points are evenly spread on either side of the line. Looks good.
2.  The Location-Scale Plot (this is displayed by default in base R's diagnostic plots). Here we're looking at the red line. If that line is more or less horizontal, then the equality of variance assumption is met.

Here all three methods agree that there aren't any issues with equality of variance:

-   the Bartlett test p-value is large and non-significant
-   the spread of points in all three groups in the residuals vs fitted graph are roughly the same
-   the red line in the scale-location graph is pretty horizontal

Overall, this assumption is pretty well met.

## Python

**Normality**

We'll be really thorough here and consider the normality of each group separately and jointly using the Shapiro-Wilk test, as well as looking at the Q-Q plot. In reality, and after these examples , we'll only use the Q-Q plot.

First, we perform the Shapiro-Wilk test on the individual groups:

```{python}
# Shapiro-Wilk on lobster groups
pg.normality(dv = "weight",
             group = "diet",
             data = lobsters_py)
```

`Flakes` and `Mussels` are fine, but, as we suspected from earlier, `Pellets` appears to have a marginally significant Normality test result.

Let's look at the Shapiro-Wilk test for all of the data together:

```{python}
# create a linear model
model = smf.ols(formula = "weight ~ C(diet)", data = lobsters_py)
# and get the fitted parameters of the model
lm_lobsters_py = model.fit()
```

```{python}
# get the residuals from the model fit
# and perform Shapiro-Wilk test
pg.normality(lm_lobsters_py.resid)
```

This on the other hand says that everything is fine. Let's look at Q-Q plot.

```{python}
#| results: hide
# Q-Q plots
p = (ggplot(lobsters_py,
        aes(sample = "weight")) +
     stat_qq() +
     stat_qq_line(colour = "red"))

p.show()
```

The Q-Q plot looks OK, not perfect, but more than good enough for us to have confidence in the normality of the data.

Overall, I'd be happy that the assumption of normality has been adequately well met here. The suggested lack of normality in the `Pellets` was only just significant and we have to take into account that there are only 5 data points in that group. If there had been a lot more points in that group, or if the Q-Q plot was considerably worse then I wouldn't be confident.

**Equality of Variance**

We'll consider the Bartlett test and we'll look at the diagnostic plots too.

```{python}
pg.homoscedasticity(dv = "weight",
                    group = "diet",
                    method = "bartlett",
                    data = lobsters_py)
```

```{python}
dgplots(lm_lobsters_py)
```

We'll just focus on the following:

1.  The Residuals Plot. What we're looking for there is that the points are evenly spread on either side of the line. Looks good.
2.  The Location-Scale Plot (this is displayed by default in base R's diagnostic plots). Here we're looking at the red line. If that line is more or less horizontal, then the equality of variance assumption is met.

Here all three methods agree that there aren't any issues with equality of variance:

-   the Bartlett test p-value is large and non-significant
-   the spread of points in all three groups in the residuals vs fitted graph are roughly the same
-   the red line in the scale-location graph is pretty horizontal

Overall, this assumption is pretty well met.
:::

#### Carry out one-way ANOVA

With our assumptions of normality and equality of variance met we can be confident that a one-way ANOVA is an appropriate test.

::: {.panel-tabset group="language"}
## R

```{r}
anova(lm_lobsters)
```

## Python

```{python}
pg.anova(dv = "weight",
         between = "diet",
         data = lobsters_py,
         detailed = True)
```
:::

> A one-way ANOVA test indicated that the mean weight of juvenile lobsters did not differ significantly between diets (p = 0.23).

#### Post-hoc testing with Tukey

In this case we did not find any significant differences between the different diets. So that is a good time for me to reiterate that carrying out the post-hoc test after getting a non-significant result with ANOVA is something that you have to think very carefully about and it all depends on what your research question it.

If your research question was:

> Does diet affect lobster weight?

or

> Is there any effect of diet on lobster weight?

Then when we got the non-significant result from the ANOVA test we should have just stopped there as we have our answer. Going digging for "significant" results by running more tests is a main factor that contributes towards lack of reproducibility in research.

If on the other hand your research question was:

> Are any specific diets better or worse for lobster weight than others?

Then we should probably have just skipped the one-way ANOVA test entirely and just jumped straight in with the Tukey's range test. The important point here is that the result of the one-way ANOVA test doesn't stop you from carrying out the Tukey test - but it's up to you to decide whether it is sensible.
:::
:::

## Summary

::: {.callout-tip}
#### Key points

-   We use an ANOVA to test if there is a difference in means between multiple continuous response variables
-   We check assumptions with diagnostic plots and check if the residuals are normally distributed
-   We use post-hoc testing to check for significant differences between the group means, for example using Tukey's range test
:::
