---
title: "Correlations"
---

```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python}
#| echo: false
#| message: false
import shutup;shutup.please()
exec(open('setup_files/setup.py').read())
```

::: callout-tip
## Learning outcomes

**Questions**

-   What are correlation coefficients?
-   What kind of correlation coefficients are there and when do I use them?

**Objectives**

-   Be able to calculate correlation coefficients in R or Python
-   Use visual tools to explore correlations between variables
-   Know the limitations of correlation coefficients
:::

## Libraries and functions

::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## tidyverse

### Libraries

```{r}
#| eval: false
# A collection of R packages designed for data science
library(tidyverse)
```

### Functions

```{r}
#| eval: false
#| warning: false
# Computes the absolute value
base::abs()

# Creates a matrix of scatter plots
graphics::pairs()

# Computes a correlation matrix
stats::cor()

# Turns object into tibble
tibble::as.tibble()

# Lengthens the data
tidyr::pivot_longer()
```

## R

### Libraries
None.

### Functions

```{r}
#| eval: false
#| warning: false
# Computes the absolute value
base::abs()

# Creates a contingency table
base::table()

# Determines the minimum value
base::which.min()

# Determines the maximum value
base::which.max()

# Creates a matrix of scatter plots
graphics::pairs()

# Computes a correlation matrix
stats::cor()

# Creates a heat map
stats::heatmap()
```


## Python

### Libraries

```{python}
#| eval: false
# A Python data analysis and manipulation tool
import pandas as pd

# Python equivalent of `ggplot2`
from plotnine import *
```

### Functions

```{python}
#| eval: false
# Compute pairwise correlation of columns
pandas.DataFrame.corr()

# Plots the first few rows of a DataFrame
pandas.DataFrame.head()

# Query the columns of a DataFrame with a boolean expression
pandas.DataFrame.query()

# Set the name of the axis for the index or columns
pandas.DataFrame.rename_axis()

# Unpivot a DataFrame from wide to long format
pandas.melt()

# Reads in a .csv file
pandas.read_csv()
```
:::
:::

## Purpose and aim

Correlation refers to the relationship of two variables (or data sets) to one another. Two data sets are said to be correlated if they are not independent from one another. Correlations can be useful because they can indicate if a predictive relationship may exist. However just because two data sets are correlated does not mean that they are causally related.

## Data and hypotheses

We will use the `USArrests` data set for this example. This rather bleak data set contains statistics in arrests per 100,000 residents for assault, murder and robbery in each of the 50 US states in 1973, alongside the proportion of the population who lived in urban areas at that time. `USArrests` is a data frame with 50 observations of five variables: `state`, `murder`, `assault`, `urban_pop` and `robbery`.

We will be using these data to explore if there are correlations between these variables.

The data are stored in the file `data/CS3-usarrests.csv`.

## Summarise and visualise

First, we load the data:

::: {.panel-tabset group="language"}
## tidyverse

```{r}
#| message: false
# load the data
USArrests <- read_csv("data/CS3-usarrests.csv")

# have a look at the data
USArrests
```

We can create a visual overview of the potential correlations that might exist between the variables. There are different ways of doing this, for example by creating scatter plots between variable pairs:

```{r}
# murder vs robbery
ggplot(USArrests,
       aes(x = murder, y = robbery)) +
    geom_point()

# assault vs urban_pop
ggplot(USArrests,
       aes(x = assault, y = urban_pop)) +
    geom_point()
```

This gets a bit tedious if there are many unique variable pairs. Unfortunately `ggplot()` does not have a pairwise function, but we can borrow the one from base R. The `pairs()` function only wants numbers, so we need to remove the `state` column for this. The `pairs()` function has a `lower.panel` argument that allows you to remove duplicate combinations (after all `murder` vs `assault` is the same as `assault` vs `murder`):

```{r}
USArrests %>% 
    select(-state) %>% 
    pairs(lower.panel = NULL)
```

## R

First, we load the data:

```{r}
# load the data
USArrests_r <- read.csv("data/CS3-usarrests.csv")

# and have a look at the data
head(USArrests_r)
```

We can only calculate correlations between numerical variables, so we have to deselect the `state` variable.

We can load the data with an extra argument, `row.names = 1`. This will instruct R to load the data but use first column (`state`) as row names:

```{r}
# load the data
USArrests_r <- read.csv("data/CS3-usarrests.csv", row.names = 1)

# have a look at the data
head(USArrests_r)
```

We can visualise the data with the `pairs()` function. This function creates a matrix of scatter plots that we can use to look for correlations. Every combination of variables appears twice (e.g. `murder` vs `assault` is the same as `assault` vs `murder`), so we use the `lower.panel = NULL` argument to only visualise the unique combinations.

```{r}
# create matrix of scatter plots
pairs(USArrests_r, lower.panel = NULL)
```

## Python
First, we load the data:

```{python}
USArrests_py = pd.read_csv("data/CS3-usarrests.csv")

USArrests_py.head()
```

We can create a visual overview of the potential correlations that might exist between the variables. There are different ways of doing this, for example by creating scatter plots between variable pairs:

```{python}
#| results: hide
# murder vs robbery
(ggplot(USArrests_py,
       aes(x = "murder",
           y = "robbery")) +
     geom_point())

# assault vs urban_pop
(ggplot(USArrests_py,
       aes(x = "assault",
           y = "urban_pop")) +
     geom_point())
```

This gets a bit tedious if there are many unique variable pairs. There is an option to automatically create a matrix of scatter plots, using Seaborn. But that would involve installing the `seaborn` package just for this. And frankly, I don't want to - not least because staring at tons of scatter plots is probably not the best way forward anyway!

If you have your heart set on creating a pairplot, then have a look [at the seaborn documentation](https://seaborn.pydata.org/generated/seaborn.pairplot.html).

:::

From the visual inspection we can see that there appears to be a slight positive correlation between all pairs of variables, although this may be very weak in some cases (`murder` and `urban_pop` for example).

## Correlation coefficients {#correlation-coefficients}

Instead of visualising the variables against each other in a scatter plot, we can also calculate correlation coefficients for each variable pair. There are different types of correlation coefficients, but the most well-known one is probably **Pearson's r**. This is a measure of the linear correlation between two variables. It has a value between -1 and +1, where +1 means a perfect positive correlation, -1 means a perfect negative correlation and 0 means no correlation at all.

There are other correlation coefficients, most notably the **Spearman's rank correlation coefficient**, a non-parametric measure of rank correlation and is generally less sensitive to outliers.

So, let's calculate **Pearson's r** for our data:

::: {.panel-tabset group="language"}
## tidyverse
We can do this using the `cor()` function. Since we can only calculate correlations between numbers, we have to remove the `state` column from our data before calculating the correlations:

```{r}
USArrests %>% 
    select(-state) %>% 
    cor()
```

This gives us a numerical overview of the Pearson's r correlation coefficients between each variable pair. Note that across the diagonal the correlation coefficients are `1` - this should make sense since, for example, `murder` is perfectly correlated with itself.

As before, the values are mirrored across the diagonal, since the correlation between, for example, `murder` and `assault` is the same as the one between `assault` and `murder`.

### Visualise the correlation matrix
Just staring at a matrix of numbers might not be very useful. It would be good to create some sort of heatmap of the values, so we can visually inspect the data a bit better. There are dedicated packages that allow you to do this (for example the [corrr](https://corrr.tidymodels.org)) package), but we'll just use `ggplot` here.

Before we can plot the data we need to reformat the data. We're taking the following steps:

1. we calculate the correlation matrix with `cor()` using the (default) method of `method = "pearson"`
2. convert the output to a tibble so we can use
3. `pivot_longer()` to reformat the data into pairwise variables and a column with the Pearson's r value
4. use the `mutate()` and `round()` functions to round the Pearson's r values

```{r}
USArrests_pear <- USArrests %>% 
    select(-state) %>% 
    cor(method = "pearson") %>% 
    as_tibble(rownames = "var1") %>% 
    pivot_longer(cols = -var1,
                 names_to = "var2",
                 values_to = "pearson_cor") %>% 
    mutate(pearson_cor = round(pearson_cor, digits = 3))
```

The output of that looks like this:

```{r}
head(USArrests_pear)
```

After all that, we can visualise the data with `geom_tile()`, adding the Pearson's r values as text labels:
```{r}
ggplot(USArrests_pear,
       aes(x = var1, y = var2, fill = pearson_cor)) +
    geom_tile() +
    geom_text(aes(label = pearson_cor),
              color = "white",
              size = 4)
```

::: {.callout-note collapse=true}
## Alternative method with rstatix
As always, there are multiple ways to skin a proverbial cat. If you'd rather use a function from the `rstatix` package (which we've loaded before), then you can run the following code, which uses the `rstatix::cor_test()` function:
```{r}
USArrests %>% 
    select(-state) %>% 
    cor_test() %>%
    select(var1, var2, cor) %>% 
    ggplot(aes(x = var1, y = var2, fill = cor)) +
    geom_tile() +
    geom_text(aes(label = cor),
              color = "white",
              size = 4)
```
:::

## R

We can do this using the `cor()` function. We can only calculate correlations between numbers, but we've already moved the `state` column from our data before calculating the correlations:

```{r}
cor(USArrests_r, method = "pearson")
```

-   The first argument is a matrix or a data frame
-   The argument `method` tells R which correlation coefficient to use (`pearson` (default), `kendall`, or `spearman`)

This gives us a numerical overview of the Pearson's r correlation coefficients between each variable pair. Note that across the diagonal the correlation coefficients are `1` - this should make sense since, for example, `murder` is perfectly correlated with itself.

As before, the values are mirrored across the diagonal, since the correlation between, for example, `murder` and `assault` is the same as the one between `assault` and `murder`.

### Visualise the correlation matrix

Just staring at a matrix of numbers might not be very useful. It would be good to create some sort of heatmap of the values, so we can visually inspect the data a bit better.

Let's save the Pearson's r correlation matrix first.
```{r}
USArrests_pear_r <- cor(USArrests_r, method = "pearson")
```

Next, we can visualise the matrix using the `heatmap()` function. This function has a lot of features that we don't need and so I'm not going to go into it in detail here. The main reason I'm using it is that it displays matrices the right way round (other plotting functions display matrices rotated by 90 degrees) and automatically labels the rows and columns.

```{r}
heatmap(USArrests_pear_r, symm = TRUE, Rowv = NA)
```

The `symm` argument tells the function that we have a symmetric matrix and in conjunction with the `Rowv = NA` argument stops the plot from reordering the rows and columns. The `Rowv = NA` argument also stops the function from adding dendrograms to the margins of the plot.

The plot itself is coloured from yellow, indicating the smallest values (which in this case correspond to no difference in correlation coefficients), through orange to dark red, indicating the biggest values (which in this case correspond to the variables with the biggest difference in correlation coefficients).

The plot is symmetric along the leading diagonal (hopefully for obvious reasons).

## Python

We can do this using the `pandas.DataFrame.corr()` function. This function takes the default `method = "pearson"` and ignores any non-numerical columns (such as the `state` column in our data set).

```{python}
USArrests_py.corr()
```
This gives us a numerical overview of the Pearson's r correlation coefficients between each variable pair. Note that across the diagonal the correlation coefficients are `1` - this should make sense since, for example, `murder` is perfectly correlated with itself.

As before, the values are mirrored across the diagonal, since the correlation between, for example, `murder` and `assault` is the same as the one between `assault` and `murder`.

### Visualise the correlation matrix
Just staring at a matrix of numbers might not be very useful. It would be good to create some sort of heatmap of the values, so we can visually inspect the data a bit better.

```{python}
# create correlation matrix
USArrests_cor_py = USArrests_py.corr()
# put the row names into a column
USArrests_cor_py = USArrests_cor_py.rename_axis("var1").reset_index()

USArrests_cor_py.head()
```

Now that we have the correlation matrix in a workable format, we need to restructure it so that we can plot the data. For this, we need to create a "long" format, using the `melt()` function.
```{python}
USArrests_pear_py = pd.melt(USArrests_cor_py,
        id_vars=['var1'],
        value_vars=['murder', 'assault', 'urban_pop', 'robbery'],
        var_name='var2',
        value_name='cor').round(3)
```

Have a look at the structure:
```{python}
USArrests_pear_py.head()
```

```{python}
#| results: hide
(ggplot(USArrests_pear_py,
        aes(x = "var1", y = "var2", fill = "cor")) +
     geom_tile() +
     geom_text(aes(label = "cor"),
               colour = "white",
               size = 10))
```
:::

The correlation matrix and visualisations give us the insight that we need. The most correlated variables are `murder` and `assault` with an $r$ value of 0.80. This appears to agree well with the set plots that we produced earlier.

## Exercise: Pearson's r for USA state data

Pearson's correlation for USA state data

We will again use the data from the file `data/CS3-statedata.csv` data set for this exercise. The data set contains 50 rows and 8 columns, with column names: `population`, `income`, `illiteracy`, `life_exp`, `murder`, `hs_grad`, `frost` and `area`.

Visually identify 3 different pairs of variables that appear to be

1.  the most positively correlated
2.  the most negatively correlated
3.  not correlated at all

Calculate Pearson's r for all variable pairs and see how well you were able to identify correlation visually.

::: {.callout-tip collapse="true"}
## Answer

Visually determining the most negative/positively and uncorrelated pairs of variables:

::: {.panel-tabset group="language"}
## tidyverse

```{r}
#| message: false
#| warning: false
USAstate <- read_csv("data/CS3-statedata.csv")

# have a look at the data
USAstate
```

The code below may look a bit daunting, but it basically repeats what we've done previously:

```{r}
USAstate_pear_cor <- USAstate %>%
    # unselect the state column
    select(-state) %>% 
    # calculate pearson's r
    cor(method = "pearson") %>%
    # convert to a tibble
    # with rownames in a column called var1
    as_tibble(rownames = "var1") %>% 
    # pivot the table
    pivot_longer(cols = -var1,
                 names_to = "var2",
                 values_to = "cor") %>% 
    # round the Pearson's r values
    mutate(cor = round(cor, digits = 3))

head(USAstate_pear_cor)
```

Next, we can plot the data:
```{r}
ggplot(USAstate_pear_cor,
       aes(x = var1, y = var2, fill = cor)) +
    # in a tile plot
    geom_tile() +
    # and add the values to the plot
    geom_text(aes(label = cor),
              color = "white",
              size = 4)
```


## R

```{r}
USAstate_r <- read.csv("data/CS3-statedata.csv",
                     row.names = 1)

# have a look at the data
head(USAstate_r)
```

```{r}
USAstate_pear_r <- cor(USAstate_r, method = "pearson")
```

```{r}
heatmap(USAstate_pear_r, symm = TRUE, Rowv = NA)
```

## Python
First, we load the data:

```{python}
USAstate_py = pd.read_csv("data/CS3-statedata.csv")

USAstate_py.head()
```

```{python}
# create correlation matrix
USAstate_cor_py = USAstate_py.corr()
# put the row names into a column
USAstate_cor_py = USAstate_cor_py.rename_axis("var1").reset_index()

USAstate_cor_py.head()
```

Now that we have the correlation matrix in a workable format, we need to restructure it so that we can plot the data. For this, we need to create a "long" format, using the `melt()` function. Note that we're not setting the `values_var` argument. If not set, then it uses all but the `id_vars` column (which in our case is a good thing, since we don't want to manually specify lots of column names).
```{python}
USAstate_pear_py = pd.melt(USAstate_cor_py,
        id_vars=['var1'],
        var_name='var2',
        value_name='cor').round(3)
```

Have a look at the structure:
```{python}
USArrests_pear_py.head()
```

```{python}
#| results: hide
(ggplot(USAstate_pear_py,
        aes(x = "var1", y = "var2", fill = "cor")) +
     geom_tile() +
     geom_text(aes(label = "cor"),
               colour = "white",
               size = 10))
```
:::

It looks like:

1.  `illiteracy` and `murder` are the most positively correlated pair
2.  `life_exp` and `murder` are the most negatively correlated pair
3.  `population` and `area` are the least correlated pair

We can explore that numerically, by doing the following:

::: {.panel-tabset group="language"}
## tidyverse

We take the correlation matrix in the long format and extract the values we're interested in:
```{r}
# most positively correlated pair
# but not 1 (self-pairs)
USAstate_pear_cor %>% 
    filter(cor != 1) %>% 
    filter(cor == max(cor))

# most negatively correlated pair
USAstate_pear_cor %>% 
    filter(cor == min(cor))

# least correlated pair
USAstate_pear_cor %>% 
    filter(cor == min(abs(cor)))
```

Note that we use the minimum *absolute* value (with the `abs()` function) to find the least correlated pair.

## R

First, we need to create the pairwise comparisons, with the relevant Pearson's $r$ values:

```{r}
# create the correlation matrix
USAstate_r_cor <- cor(USAstate_r, method = "pearson")

# build a contingency table with as.table()
# and create a dataframe with as.data.frame()
USAstate_r_cor <- as.data.frame(as.table(USAstate_r_cor))
    
# and have a look
head(USAstate_r_cor)
```

Is this method obvious? No! Some creative Googling led to [Stackoverflow](https://stackoverflow.com/questions/7074246/show-correlations-as-an-ordered-list-not-as-a-large-matrix) and here we are. But, it does give us what we need.

Now that we have the paired comparisons, we can extract the relevant data:

```{r}
# first we remove the same-pair correlations
USAstate_r_cor <- subset(USAstate_r_cor, Freq != 1)

# most positively correlated pair
USAstate_r_cor[which.max(USAstate_r_cor$Freq), ]

# most negatively correlated pair
USAstate_r_cor[which.min(USAstate_r_cor$Freq), ]

# least correlated pair
USAstate_r_cor[which.min(abs(USAstate_r_cor$Freq)), ]
```

Note that we use the minimum *absolute* value (with the `abs()` function) to find the least correlated pair.

## Python
We take the correlation matrix in the long format:
```{python}
USAstate_pear_py.head()
```

and use it to extract the relevant values:
```{python}
# filter out self-pairs
df_cor = USAstate_pear_py.query("cor != 1")

# filter for the maximum correlation value
df_cor[df_cor.cor == df_cor.cor.max()]

# filter for the minimum correlation value
df_cor[df_cor.cor == df_cor.cor.min()]

# filter for the least correlated value
# create a column containing absolute values
df_cor["abs_cor"] = df_cor["cor"].abs()
df_cor[df_cor.abs_cor == df_cor.abs_cor.min()]
```
:::
:::

## Spearman's rank correlation coefficient

This test first calculates the rank of the numerical data (i.e. their position from smallest (or most negative) to the largest (or most positive)) in the two variables and then calculates Pearson's product moment correlation coefficient using the ranks. As a consequence, this test is less sensitive to outliers in the distribution.

::: {.panel-tabset group="language"}
## tidyverse

```{r}
USArrests %>% 
    select(-state) %>% 
    cor(method = "spearman")
```

## R

```{r}
cor(USArrests_r,
    method = "spearman")
```

## Python
```{python}
USArrests_py.corr(method = "spearman")
```
:::

## Exercise: State data (Spearman)

Spearman's correlation for USA state data

Calculate Spearman's correlation coefficient for the `data/CS3-statedata.csv` data set.

Which variable's correlations are affected most by the use of the Spearman's rank compared with Pearson's r? Hint: think of a way to address this question programmatically.

Thinking about the variables, can you explain why this might this be?

::: {.callout-tip collapse="true"}
## Answer
In order to determine which variables are most affected by the choice of Spearman vs Pearson you could just plot both matrices out side by side and try to spot what was going on, but one of the reasons we're using programming languages is that we can be a bit more **programmatic** about these things. Also, our eyes aren't that good at processing and parsing this sort of information display. A better way would be to somehow visualise the data.

::: {.panel-tabset group="language"}
## tidyverse
First, calculate the Pearson and Spearman correlation matrices (technically, we've done the Pearson one already, but we're doing it again for clarity here).
```{r}
cor_pear <- USAstate %>% 
    select(-state) %>% 
    cor(method = "pearson")

cor_spear <- USAstate %>% 
    select(-state) %>% 
    cor(method = "spearman")
```

We can calculate the difference between two matrices by subtracting them.
```{r}
cor_diff <- cor_pear - cor_spear
```

Again, we could now just look at a grid of 64 numbers and see if we can spot the biggest differences, but our eyes aren't that good at processing and parsing this sort of information display. A better way would be to visualise the data.

```{r}
cor_diff <- cor_diff %>% 
    # convert to a tibble
    # with rownames in a column called var1
    as_tibble(rownames = "var1") %>% 
    # pivot the table
    pivot_longer(cols = -var1,
                 names_to = "var2",
                 values_to = "cor") %>% 
    # round the correlation values
    mutate(cor = round(cor, digits = 3))

head(cor_diff)
```

We're interested in the magnitude of difference, so we're looking for the maximum absolute difference. So we can add an extra column that contains these values.

```{r}
cor_diff <- cor_diff %>% 
    mutate(abs_cor = abs(cor))
```

Next, we can plot the data:
```{r}
ggplot(cor_diff,
       aes(x = var1, y = var2, fill = abs_cor)) +
    # in a tile plot
    geom_tile() +
    # and add the values to the plot
    geom_text(aes(label = abs_cor),
              color = "white",
              size = 4)
```

## R
First, calculate the Pearson and Spearman correlation matrices (technically, we've done the Pearson one already, but we're doing it again for clarity here).

```{r}
cor_pear_r <- cor(USAstate_r, method = "pearson")
cor_spea_r <- cor(USAstate_r, method = "spearman")
```

We can calculate the difference between two matrices by subtracting them.
```{r}
cor_diff_r <- cor_pear_r - cor_spea_r
```

Again, we could now just look at a grid of 64 numbers and see if we can spot the biggest differences, but our eyes aren't that good at processing and parsing this sort of information display. A better way would be to visualise the data.

```{r}
heatmap(abs(cor_diff_r), symm = TRUE, Rowv = NA)
```

The `abs()` function calculates the absolute value (i.e. just the magnitude) of the matrix values. This is just because I only care about situations where the two correlation coefficients are different from each other but I don't care which is the larger. The `symm` argument tells the function that we have a symmetric matrix and in conjunction with the `Rowv = NA` argument stops the plot from reordering the rows and columns. The `Rowv = NA` argument also stops the function from adding dendrograms to the margins of the plot.

## Python
First, calculate the Pearson and Spearman correlation matrices (technically, we've done the Pearson one already, but we're doing it again for clarity here).

```{python}
cor_pear_py = USAstate_py.corr(method = "pearson")
cor_spea_py = USAstate_py.corr(method = "spearman")
```

We can calculate the difference between two matrices by subtracting them.
```{python}
cor_dif_py = cor_pear_py - cor_spea_py
```

Again, we could now just look at a grid of 64 numbers and see if we can spot the biggest differences, but our eyes aren't that good at processing and parsing this sort of information display. A better way would be to visualise the data.

```{python}
# get the row names in a column
cor_dif_py = cor_dif_py.rename_axis("var1").reset_index()

# reformat the data into a long format
# and round the values
cor_dif_py = pd.melt(cor_dif_py,
        id_vars=['var1'],
        var_name='var2',
        value_name='cor').round(3)
        
# create a column with absolute correlation difference values
cor_dif_py["abs_cor"] = cor_dif_py["cor"].abs()

# have a look at the final data frame
cor_dif_py.head()
```

Now we can plot the data:
```{python}
#| results: hide
(ggplot(cor_dif_py,
        aes(x = "var1", y = "var2", fill = "abs_cor")) +
     geom_tile() +
     geom_text(aes(label = "abs_cor"),
               colour = "white",
               size = 10))
```
:::

All in all there is not a huge difference in correlation coefficients, since the values are all quite small. Most of the changes occur along the `area` variable. One possible explanation could be that certain states with a large area have a relatively large effect on the Pearson's r coefficient. For example, Alaska has an area that is over twice as big as the next state - Texas.

If, for example, we'd look a bit closer then we would find for `area` and `income` that Pearson gives a value of 0.36, a slight positive correlation, whereas Spearman gives a value of 0.057, basically uncorrelated.

This means that this is basically ignored by Spearman.

Well done, [Mr. Spearman](https://en.wikipedia.org/wiki/Charles_Spearman).
:::

## Key points

::: callout-note
-   Correlation is the degree to which two variables are linearly related
-   Correlation does not imply causation
-   We can visualise correlations by plotting variables against each other or creating heatmap-type plots of the correlation coefficients
-   Two main correlation coefficients are Pearson's r and Spearman's rank, with Spearman's rank being less sensitive to outliers
:::
